\chapter{Implementation}

\section{Django Stack Development}

\subsection{Backend Setup}

Development used Windows 11, Python 3.12, virtual environment, and Git. Local setup:

\begin{verbatim}
python -m venv venv
pip install -r requirements.txt
python manage.py migrate
python manage.py loaddata fixtures/initial_data.json
python create_admin.py
python manage.py runserver
\end{verbatim}

\subsection{Frontend Setup}

Templates live under \texttt{templates/}; static assets under \texttt{static/}. \texttt{collectstatic} prepares files for production; WhiteNoise serves them on Render.

\section{Configuration Setup}

\subsection{Middleware}

Order: SecurityMiddleware, WhiteNoise, Session, Common, CSRF, Authentication, Messages, XFrameOptions.

\subsection{Routes}

URL configuration in \texttt{docbook/urls.py} includes apps for accounts, patients, doctors, bookings, core, and admin views.

\subsection{Database Models}

Key models: \texttt{User}, \texttt{Profile}, \texttt{Booking}, \texttt{Prescription}, \texttt{Review}, day schedule models, and \texttt{TimeRange}.

\section{Authentication and Authorization}

Login uses Django \texttt{authenticate()} and session cookies. Passwords are hashed with PBKDF2-SHA256. RBAC mixins guard class-based views; decorators guard selected function views.

\section{Appointment Booking Workflow}

\subsection{Schedule and Slot Generation}
Doctors configure weekly availability; the UI derives bookable slots from time ranges minus existing bookings.

\subsection{Booking Creation and Status Management}
Patients create bookings with status \texttt{pending}; doctors confirm, complete, cancel, or mark \texttt{no\_show}.

\subsection{Prescription and Review Processing}
Doctors add prescriptions via CKEditor; patients submit reviews after completed visits.

\section{Administrative Dashboard and Reports}

Custom admin at \texttt{/admin/} provides statistics, user lists, speciality CRUD, and appointment/revenue reports. Django admin remains at \texttt{/super-admin/}.

\section{Cloud Deployment on Render}

Build command: \texttt{bash build.sh}. Start command: \texttt{gunicorn docbook.wsgi:application --bind 0.0.0.0:\$PORT}. Environment variables: \texttt{SECRET\_KEY}, \texttt{DEBUG=False}, \texttt{ALLOWED\_HOSTS}, \texttt{DATABASE\_URL}, \texttt{CSRF\_TRUSTED\_ORIGINS}.

Live URL: \url{https://healthcare-appointment-system-9a2r.onrender.com}

\section{Security and HTTPS Configuration}

When \texttt{DEBUG=False}, settings enable SSL redirect, secure cookies, HSTS, and trusted CSRF origins.

\section{User Interface Implementation}

The public home page supports doctor search; role dashboards show appointments and actions appropriate to each user type.
